---
title: "STA 141A: Final Project Report"
author: "Sunetra Maddipatla"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("xgboost")
install.packages("slider")
install.packages("nnet")
install.packages("ranger")
install.packages("pROC")
```

```{r}

wildfires <- read_csv("~/Desktop/wildfires_sample_100k.csv") %>% 
  select(FIRE_YEAR, DISCOVERY_DOY, FIRE_SIZE, LATITUDE, LONGITUDE, STATE) %>%
  mutate(STATE = as.factor(STATE))

glimpse(wildfires)

```

```{r}
## Setup
suppressMessages({
    library(dplyr)
    library(ggplot2)
    library(tidyverse)
    library(slider)
    library(pROC)
    library(xgboost)
    library(Matrix)
    library(nnet)
})
```

Temporal Patterns Convert `FIRE_YEAR` and `DISCOVERY_DOY` to `FIRE_DATE`

```{r}
fires <- wildfires %>% 
    mutate(FIRE_DATE = as.Date(paste(FIRE_YEAR, DISCOVERY_DOY, sep = "-"), format = "%Y-%j"),
          YEAR = year(FIRE_DATE),
          MONTH = month(FIRE_DATE))
fires
```

Filtering raw wildfire records to retain only "notable" fires

```{r}
notable_threshold_acres <- 300
notable_fires <- fires %>%
  filter(FIRE_SIZE >= notable_threshold_acres)
notable_fire

cat("Number of notable fires:", nrow(notable_fires))
#clear deginition and justification of the notable fire cutoff
```

Spatial Patterns

Space-Time Units

```{r}
#spacetime units
# ---- Create grid IDs and month index ----

notable_fires <- notable_fires %>%
  mutate(
    lat_bin = floor(LATITUDE),
    lon_bin = floor(LONGITUDE),
    ym = YEAR * 100 + MONTH   # e.g., 201507
  )

# Fire indicator at (grid, month): did any notable fire occur?

grid_month_fire <- notable_fires %>%
  group_by(lat_bin, lon_bin, ym) %>%
  summarise(
    n_fires = n(),
    .groups = "drop"
  ) %>%
  mutate(
    fire_this_month = 1
  )

head(grid_month_fire)

```

Construction of a panel for these patterns

```{r}
# ---- Build panel (all grid cells that appear at least once) ----
all_grids <- grid_month_fire %>% distinct(lat_bin, lon_bin)

# Generate a safe monthly sequence using actual dates
start_date <- ymd(paste0(min(grid_month_fire$ym) %/% 100, "-", min(grid_month_fire$ym) %% 100, "-01"))
end_date   <- ymd(paste0(max(grid_month_fire$ym) %/% 100, "-", max(grid_month_fire$ym) %% 100, "-01"))

all_months <- tibble(date = seq.Date(start_date, end_date, by = "month")) %>%
  mutate(year = year(date), month = month(date), ym = year * 100 + month) %>%
  select(ym, year, month)

panel <- crossing(all_grids, all_months) %>%
  left_join(grid_month_fire, by = c("lat_bin", "lon_bin", "ym")) %>%
  mutate(
    n_fires = replace_na(n_fires, 0),
    fire_this_month = if_else(is.na(fire_this_month), 0, fire_this_month)
  ) %>%
  arrange(lat_bin, lon_bin, ym)

glimpse(panel)
```

Construction Of Binary Outcome Variable

```{r}
#binary otucome variable

month_fires <- notable_fire %>%
  group_by(STATE, FIRE_YEAR, FIRE_MONTH) %>%
  summarise(
    notable_fire_count = n(),
    avg_fire_size = mean(FIRE_SIZE, na.rm = TRUE),
    max_fire_size = max(FIRE_SIZE, na.rm = TRUE),
    wildfire_occurrence = 1,
    .groups = "drop"
  )
month_fires
```

```{r}
#IFFFYYYY
month_fires <- notable_fire %>%
  group_by(STATE, FIRE_YEAR, FIRE_MONTH) %>%
  summarise(
    notable_fire_count = n(),
    avg_fire_size = mean(FIRE_SIZE, na.rm = TRUE),
    max_fire_size = max(FIRE_SIZE, na.rm = TRUE),
    wildfire_occurrence = ifelse(notable_fire_count > 0, 1, 0),
    .groups = "drop"
  )
```

///////

Prediction Target

```{r}
panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>%
  mutate(
    y_next = lead(fire_this_month, n = 1L)
  ) %>%
  ungroup() %>%
  filter(!is.na(y_next))

table(panel$y_next)
```

Baseline Predictor

```{r}
mode_state <- function(x) {
  names(sort(table(x), decreasing = TRUE))[1]
}

grid_state <- notable_fires %>%
  group_by(lat_bin, lon_bin) %>%
  summarise(
    state = mode_state(STATE),
    .groups = "drop"
  ) %>%
  mutate(state = as.factor(state))
panel <- panel %>%
  left_join(grid_state, by = c("lat_bin","lon_bin"))

panel %>% select(`state`) %>% summary

```

Seasonality: Cyclical month encoding

```{r}
panel <- panel %>%
  mutate(
    month_sin = sin(2 * pi * month / 12),
    month_cos = cos(2 * pi * month / 12)
  )

panel %>% select('month_sin', 'month_cos') %>% head
```

Lagged fire history

```{r}
grid_fire_size <- notable_fires %>%
  group_by(lat_bin, lon_bin, ym) %>%
  summarise(
    fire_size = mean(FIRE_SIZE, na.rm = TRUE),
    .groups = "drop"
  )


panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>%
  mutate(
    fire_lag12 = lag(n_fires, 12, default = 0),
    fire_roll3 = lag(n_fires, 1, default = 0) +
                 lag(n_fires, 2, default = 0) +
                 lag(n_fires, 3, default = 0),
    fire_roll12 = stats::filter(lag(n_fires, default = 0),
                                rep(1, 12),
                                sides = 1) |> as.numeric(),
    size_roll3 = lag(fire_size, 1, default = 0) +
                 lag(fire_size, 2, default = 0) +
                 lag(fire_size, 3, default = 0),
    size_roll12 = stats::filter(lag(fire_size, default = 0),
                                rep(1, 12),
                                sides = 1) |> as.numeric()
  ) %>%
  ungroup() %>%
  mutate(
    fire_roll12 = coalesce(fire_roll12, 0),
    size_roll12 = coalesce(size_roll12, 0)
  )

panel %>% 
  select("n_fires", "fire_lag12", "fire_roll3", "fire_roll12", 
         "size_roll3", "size_roll12") %>%
  summary
```

Dataset Summary Table

```{r}
#dataset summary table
dataset_summary <- tibble(
  n_rows = nrow(panel),
  year_min = min(panel$year),
  year_max = max(panel$year),
  spatial_unit = "1° x 1° lat-lon grid (floor bins)",
  temporal_unit = "monthly",
  outcome = "y_next = 1 if any notable fire occurs in grid cell next month",
  predictors = "lat_bin, lon_bin, state, month_cos, month_sin, year, 
    n_fires, fire_lag12, fire_roll3, fire_roll12, 
    size_roll3, size_roll12"
)

dataset_summary
```

## 2 Explanatory Data Analysis 

EDA Plot 1: Spatial pattern

```{r}

eda_spatial <- panel %>%
  group_by(lat_bin, lon_bin) %>%
  summarise(
    fire_rate = mean(y_next),
    .groups = "drop"
  )

ggplot(eda_spatial, aes(x = lon_bin, y = lat_bin, fill = fire_rate)) +
  coord_quickmap(xlim = c(-125, -65), ylim = c(25, 50)) + 
  geom_tile() +
  labs(
    title = "Spatial pattern: Mean next-month notable-fire rate by grid cell",
    x = "Longitude bin (floor)",
    y = "Latitude bin (floor)",
    fill = "Mean y_next"
  )
```

## EDA Plot 2: Temporal pattern

```{r}
eda_month <- panel %>%
  group_by(month) %>%
  summarise(
    mean_y_next = mean(y_next),
    .groups = "drop"
  )

ggplot(eda_month, aes(x = month, y = mean_y_next)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Seasonality: Mean probability of a notable fire in the next month",
    x = "Month",
    y = "Mean of y_next"
  )
```

## 3) Modeling

```{r}
model_df <- panel %>%
  mutate(y_next = factor(y_next, levels = c(0, 1))) %>%
  select(y_next, lat_bin, lon_bin, state, 
         month_cos, month_sin, year, 
         n_fires, fire_lag12, fire_roll3, fire_roll12, size_roll3, size_roll12)
glimpse(model_df)
```

Gradient Boosted Trees (XGBoost)
```{r}
library(xgboost)

xg_df <- model_df %>%
  mutate(y_next_num = factor(y_next, levels = c(0, 1)))

x_mat <- model.matrix(
  y_next_num ~ lat_bin + lon_bin + state +
    month_cos + month_sin + year +
    n_fires + fire_lag12 + fire_roll3 + fire_roll12 +
    size_roll3 + size_roll12,
  data = xg_df
)[, -1]

xg_fit <- xgboost(
  x = x_mat,
  y = xg_df$y_next_num,
  objective = "binary:logistic",
  eval_metric = "auc",
  nrounds = 100,
  max_depth = 4,
  learning_rate = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8,
  verbosity = 0
)

xg_fit
```
```{r}
library(ranger)

rf_model <- ranger(
  y_next ~ lat_bin + lon_bin + state +
    month_cos + month_sin + year +
    n_fires + fire_lag12 + fire_roll3 + fire_roll12 +
    size_roll3 + size_roll12,
  data = model_df,
  probability = TRUE,
  num.trees = 500
)

rf_model
```

## 4) Model Fitting and Validation

```{r}
train <- model_df %>% filter(year <= years_sorted[5])
valid <- model_df %>% filter(year == years_sorted[6])

nn_fit <- nnet(
  feat_formula,
  data = train,
  size = 5,
  decay = 0.01,
  maxit = 200,
  trace = FALSE
)

pred_test <- predict(nn_fit, newdata = valid, type = "raw")

head(pred_test)
colnames(pred_test)
```


```{r}
years_sorted <- sort(unique(model_df$year))
min_train_years <- 3

feat_formula <- y_next ~ lat_bin + lon_bin + month_cos + month_sin + fire_roll12
features <- c("month_cos", "month_sin", "lat_bin", "lon_bin", "fire_roll12")
target <- "y_next"

fold_results <- tibble(
  train_end_year = years_sorted[min_train_years:(length(years_sorted) - 2)],
  valid_year = years_sorted[(min_train_years + 1):(length(years_sorted) - 1)]
) %>%
  mutate(
    auc_glm = map2_dbl(train_end_year, valid_year, function(tr_end, va_year) {
      train <- model_df %>% filter(year <= tr_end)
      valid <- model_df %>% filter(year == va_year)
      
      fit <- glm(feat_formula, data = train, family = binomial)
      p_hat <- predict(fit, newdata = valid, type = "response")
      
      as.numeric(auc(roc(valid$y_next, p_hat, quiet = TRUE)))
    }),
    
    auc_nn = map2_dbl(train_end_year, valid_year, function(tr_end, va_year) {
  
  train <- model_df %>% filter(year <= tr_end)
  valid <- model_df %>% filter(year == va_year)
  
  nn_fit <- nnet(
    feat_formula,
    data = train,
    size = 5,
    decay = 0.01,
    maxit = 200,
    trace = FALSE
  )
  
  p_hat <- predict(nn_fit, newdata = valid, type = "raw")
  p_hat <- as.numeric(p_hat)
  
  as.numeric(auc(roc(valid$y_next, p_hat, quiet = TRUE)))
})

)

fold_results
```

```{r}
fold_long <- fold_results %>%
  pivot_longer(
    cols = c(auc_glm, auc_nn),
    names_to = "model",
    values_to = "auc"
  )

ggplot(fold_long, aes(x = valid_year, y = auc, color = model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Time-aware cross-validation diagnostic: AUC by validation year",
    x = "Validation year",
    y = "AUC",
    color = "Model"
  ) + 
  theme_minimal()
```

```{r}
cat('Overall, Neural Network outperforms logistic regression in AUC by ',
    round(mean(fold_results$auc_nn) / mean(fold_results$auc_glm) * 100, 2) - 100, "%", sep = '')
```

##5) Summary

```{r}

```
